Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimization Choices

### 2025-03-04 jso.qmd
proofread done

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimization for the guided tour. It is of
interest to assess how quickly and closely the optimizer reaches a
global optimum, for various PP indexes that may have differing
complexities. To observe the performance of JSO with different types of
PP indexes, metrics are introduced to capture specific properties of the
index including squintability [based on @barnett1981interpreting's
squint angle] and smoothness. We mathematically define metrics for
squintability and smoothness, which is a new contribution for PP
research. A series of simulation experiments are conducted using JSO to
detect different target patterns (pipe and sine-wave) with different PP
indexes (holes, MIC, TIC, dcor, loess, splines, skinny, and stringy), as
well as JSO's hyper-parameter choices (number of jellyfish and maximum
number of tries). This work should facilitate better optimization for PP
and guide the choice of optimizer when designing new PP indexes.

@@

To solve the optimization problem embedded in the PPGT with JSO, a starting
projection, an index function, the number of jellyfish, and the maximum
number of trials (tries) are provided as input. Then, the current
projection is evaluated by the index function. The projection is then
moved in a direction determined by a random factor, influenced by how
far along we are in the optimization process. Occasionally, completely
new directions may be taken like a jellyfish might with ocean currents.
A new projection is accepted if it is an improvement compared to the
current one, rejected otherwise. This process continues and iteratively
improves the projection, until the pre-specified maximum number of
trials is reached.

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of tries.

@@

This set of simulations is expanded to assess the performance of JSO
relative to the index properties, with a second data set type and
additional PP indexes. The second data set has a sine wave in two
dimensions and Gaussian noise in the remaining dimensions. Seven
additional PP indexes that should be able to extract the sine wave are
used: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `splines2d`, `skinny` and `stringy2`,
following @laa_using_2020. (The supplementary material includes
definitions of these PP indexes, and descriptions of the pipe and sine
wave data simulation.). The JSO hyper-parameters, number of jellies and
maximum tries, are also included. This results in a total of
`r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine wave data. The
choice of 4-12 dimensions covers the range of easy to relatively
difficult for the optimizers to detect the structured projections.
<!-- A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability.-->
The same approach is used, that each experiment is repeated 50 times and success
rate is calculated at each level of the simulation design.

@@

Recall that smoothness and squintability are metrics used to characterize the index 
function for PP tasks and are invariant to the optimizer and its 
hyper-parameters. Thus simulations with different JSO hyper-parameters 
for the same PP task share the same smoothness value. In total, smoothness and
squintability metrics are calculated on the 22 PP tasks in our simulation.

@@

A generalized linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting the success rate.
Predictors are smoothness, squintability, data dimension, and JSO
hyper-parameters. <!-- The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine wave shape.-->

@@

```{r}
#| fig.height: 7.5
#| fig.width: 10
#| label: fig-proj
#| fig.align: "center"
#| fig.cap: "Visual comparison of JSO and CRS results, using optimal data projections obtained over 50 simulations of the pipe data. Rows correspond to data dimension. Columns correspond to quantiles of the index values, with 0 being minimum, 100 being maximum, and 50 the median. JSO achieves the better views of the pipe generally than CRS. As dimension increases both have more difficulty finding the pipe."

@@

The performance of JSO is assessed relative to the existing CRS
optimizer based on the optimal projection obtained. The final
projections for the pipe data sets found by the two optimizers are shown
in @fig-proj, with rows corresponding to data dimension. The 50
simulations in each data dimension are sorted by index value, and the
projections corresponding to each 10th quantile value are shown, with
minimum at left and maximum at right. Index value is printed on the
plot. The purpose is to summarize the views of the data resulting from
different optimization, and hence compare results between the two
optimizers. Generally, the JSO does a more consistent job of finding the
pipe structure clearly. As the data dimension increases, both optimizers
struggle and less clearly capture the circle shape.

<!-- broken down by 10th quantile,
faceted by the data dimensions. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The CRS also finds the pipe
shape but with a wide rim, suggesting a further polish search may be
required. With increasing dimensions, JSO may not always identify the
pipe shape due to random sampling, but it still finds the pipe shape in
over 50% of cases. Compared to CRS, JSO achieves higher index values and
clearer pipe shapes across all quantiles in data of 8, 10, 12
dimensions, suggesting its advantage in exploring high-dimensional
spaces.-->

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is summarized in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For problems with relatively lower dimensional (4,
6) search spaces, small parameter values (20 jellyfish and a maximum of
50 tries) are sufficient to achieve a high success rate. Many more
jellyfish (100 jellyfish and a maximum of 100 tries) are needed to
obtain a mediocre success rate for dimensions 10 and 12. This suggests
that even JSO will struggle with high dimensions, and require
substantial more computational time to have an acceptable success rate,
especially for difficult PP indexes based on scagnostics.
<!-- dimensions) where a more complex search is necessary. However, in lower-dimensional problems, the benefit may be less pronounced, as the space is already relatively easy to navigate. While increasing both parameters enhances the performance of JSO, it also extends the computational time required for the optimization, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.-->

```{r}
#| label: fig-proportion
#| fig.width: 8
#| fig.height: 3.5
#| fig.align: "center"
#| fig.cap: "Summary of the relationship between JSO hyper-parameters on optimization success, using the holes index for pipe data of dimensions 4-12. Bootstrap samples show the variability in success rate. Success rate mostly plateaus by 50 jellies. The JSO has some difficulty finding the pipe when dimension is higher than 8! Maximum number of tries has little effect."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

@@

res_quantile <- res_interval_df |> 
  unnest(df) |> 
  group_by(n_jellies, max_tries, case_id, d) |> 
  select(-I_max, -id) |> 
  distinct() |> 
  mutate(d = as.factor(d),
         n_jellies = ifelse(d == 10, n_jellies - 1, n_jellies),
         n_jellies = ifelse(d == 12, n_jellies + 1, n_jellies)) |> 
  rename(`max # of tries` = max_tries)

res_quantile |> 
  ungroup() |> 
  ggplot(aes(x = n_jellies - 0.5, y = proportion, 
             group = d, color = d)) + 
  geom_segment(aes(xend = n_jellies + 0.5), size = 0.05) + 
  geom_segment(aes(x = n_jellies - 1, xend = n_jellies + 1), 
               data = res_quantile |> summarise(proportion = mean(proportion)), 
               size = 2) + 
  geom_line(aes(x = n_jellies), 
               data = res_quantile |> summarise(proportion = mean(proportion)), 
            alpha = 0.3) + 
  scale_color_brewer(palette = "Dark2", name = "dimension") +
  scale_x_continuous(breaks = c(20, 50, 100), labels = c("20", "50", "100")) +
  facet_wrap(vars(`max # of tries`), 
             labeller = "label_both") + 
  theme_bw() +
  theme(panel.grid.minor = element_blank()) + 
  xlab("Number of jellyfish") + 
  ylab("Proportion of success")
```

@@

Smoothness and squintability are calculated across a collection of pipe
and sine wave finding problems to construct the relationship between
success rate, JSO hyper-parameters, and index properties.
@tbl-smoothness-squintability presents values computed for a subset of
PP indexes on the sine wave data for dimensions 4-8. (A full table of
the parameters estimated from the Gaussian process (outputscale $\eta$,
lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and from the
scaled logistic function ($\theta_1$ to $\theta_4$) for calculating
smoothness and squintability for all PP indexes and data combinations is
included in the supplementary materials). The column $\nu$ is used as
the smoothness metric and the column $\varsigma$ is calculated as
equation (@eq-squintability-parametric) as the squintability metric.

@fig-idx-proj-dist shows a sample of traces from the indexes in
@tbl-smoothness-squintability for the 6D sine wave data. The traces the
PP index values computed for projections along an interpolated path
between a noise projection and the optimal projection (sine wave). (Note
index values were self-standardized to range between 0 and 1, for
comparison.) The `splines` index is very smooth with a gradual increase
in value between the noise projection and the optimal projection. The
`TIC` index is slightly less smooth and slightly less gradual change in
value. Least desirable is the `skinny` index which is noisier and the
optimization needs to be much closer to the optimal angle to have a
higher value. This illustrates how the new measures for squintability
and smoothness work to describe the PP indexes.

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: The columns $\nu$ and $\varsigma$ show the smoothness and squintability measures for three PP indexes for the sine wave data. The splines index has the largest squint angle and is the smoothest. The skinny index has low smoothness and smallest squint angle. As dimension increases the squint angle gets smaller but smoothness doesn't change much.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 17)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor", index),
         index = ifelse(index == "loess2d", "loess", index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  filter(shape == "sine", index %in% c("TIC", "splines", "skinny")) |>
  select(shape, index, d, smooth, squint) |>
  arrange(desc(squint)) |>
  knitr::kable(
    digits = 4, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      #"$\\eta$", "$\\ell$", 
      "$\\nu$", #"$\\sigma$",
      #"$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", 
      "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rr") #align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  #kable_styling(font_size = 10) 
```

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum number of
tries, as predictors. The fit suggests that the JSO success rate is only
affected by squintability, dimension and number of jellies. As expected,
the JSO success rate is higher with higher squintability and/or more
jellyfish, and lower when dimension is higher (higher-dimensional
optimization is more difficult). Interestingly, the JSO is unaffected by
the smoothness of the index function. This is consistent with the way
random search algorithms jump from value to value without taking local
regularity into account, as opposed to gradient-based optimizers for
example. Allowing for more tries also does not affect success rate
significantly. A unit increase in squintability increases the success
rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As
dimension increases by one the success rate almost halves. Increasing
the number of jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.
@tbl-joint-output shows the fit statistics supporting that the model is
reasonably strong.

Note that the addition of interaction terms for squintability and data
dimension, and smoothness and data dimension, improves the model a small
amount, but does not change the interpretation above.

```{r}
#| tbl-pos: H
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, number of jellyfish and maximum number of tries. Interestingly, squintability and dimension strongly affect jellyfish optimization success. The number of jellies marginally affects success, but index smoothness, and increasing the number of tries do not."
#| fig-pos: t
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate = TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    #long_time = "Long runtime", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of tries")) |>
  bold_p(t = 0.1)
```

@@

<!--The simulation result suggests that squintability and smoothness are likely dependent. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness. While it may seem counterintuitive at first that smoothness is not a significant factor, this outcome is reasonable in this context, as the swarm-based method used in this study, JSO, is designed to perform well even in non-smooth search spaces by leveraging population-based exploration.-->

This study intends to serve as an illustration that the proposed
measures effectively captures the key factors influencing the success
rate of PP.
<!--The interactions between the two variables were not examined in this study due to the limited number of data points, nor was the linearity assumption of the relationship with success rate explicitly tested.-->

@@

The simulation studies have compared the JSO with CRS and shown how
smoothness and squintability affect the JSO success rate. Here we
explain how they can be used for new index and optimizer development.
Using the JSO optimizer for PPGT requires a change in user behavior. For
all the existing optimizers a single optimization path is followed. The
JSO has multiple optimization paths, and needs additional tools to
incorporate into the PPGT, as explained here.

@@

To use JSO for PPGT in the `tourr` package, specify
`search_f = search_jellyfish` in the guided tour. Unlike existing
optimizers, the animation function `animate_*()` won't directly render
the tour path for JSO due to the generation of multiple tour paths. To
visualize the path visited by each individual jellyfish, assign the
animation to an object (`res <- animate_xy(...)`). This will save the
bases visited by JSO, along with relevant metadata, as a tibble data
object [see @RJ-2021-105 for more details on the data object]. The bases
visited for individual jellyfish can then be extracted and viewed using
the `planned_tour()` function.

@@

The `ferrn` package [@RJ-2021-105] provides functionality for computing
the smoothness and squintability metrics. Both metrics require sampling
random bases using the `sample_bases()` function, followed by the metric
calculation with `calc_smoothness()` or `calc_squintability()`.

@@

To sample bases for calculating smoothness, the following arguments are
required: the index function, the dataset, and the number of random
bases to sample. The output of `sample_bases()` is a data frame with a
list-column of sampled basis matrix and index value. Particularization
is available to speed up the index value computation through the
`parallel = TRUE` argument.

The `calc_smoothness()` function takes the output from `sample_bases()`
and fits a Gaussian process model to the index values against the
sampled basis, as the location, to obtain the smoothness metric.
Starting parameters and additional Gaussian process arguments can be
specified and details of the fit can be accessed through the `fit_res`
attribute of the output.

@@

Bases sampling for squintability includes an additional step of
interpolating between the sampled bases and the best basis. This step is
performed when the arguments `step_size` and `min_proj_dist` in the
`sample_bases()` function are set to non-NA numerical values. Given the
projection distance typically ranging from 0 to 2, it is recommended to
set `step_size` to 0.1 or lower, and `min_proj_dist` to be at least 0.5
to ensure a meaningful interpolation length.

The `calc_squintability()` function computes squintability using two
methods: 1) parametrically, by fitting a scaled sigmoid function through
non-linear least square (`method = "nls"`), and 2) non-parametrically,
using kernel smoothing (`method = "ks"`). A `bin_width` argument is
required to average the index values over the projection distance before
the fit. For the parametric case, the output provides the estimated
parameters for the scaled sigmoid function ($\theta_1$ to $\theta_4$)
and the calculated squintability metric as Equation
(@eq-squintability-parametric). For the non-parametric case, it shows
the maximum gradient attained (`max_d`), the corresponding projection
distance (`max_dist`), and the squintability metric as their products.

@@

This paper has presented new metrics to mathematically define desirable
features of PP indexes, squintability and smoothness, and used these to
assess the performance of the new jellyfish search optimizer. The
metrics will be generally useful for characterizing PP indexes, and help
with developing new indexes.

@@

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optimum when more jellyfish are used and
the maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optimum. However, interestingly smoothness does not affect
the JSO performance.

One future direction of this work is to test the JSO, along with its
variations and other swarm-based optimizers, on a broader range of
indexes, for example, scagnostic indexes.

>>>>

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimization for the guided tour. It is of
interest to assess how quickly and closely the optimizer reaches a
global optimum, for various PP indexes that may have differing
complexities. To observe the performance of JSO with different types of
PP indexes, metrics are introduced to capture specific properties of the
index including squintability [based on @barnett1981interpreting's
squint angle] and smoothness. We mathematically define metrics for
squintability and smoothness, which is a new contribution for PP
research. A series of simulation experiments are conducted using JSO to
detect different target patterns (pipe and sine-wave) with different PP
indexes (holes, MIC, TIC, dcor, loess, splines, skinny, and stringy), as
well as JSO's hyper-parameter choices (number of jellyfish and maximum
number of iterations). This work should facilitate better optimization for PP
and guide the choice of optimizer when designing new PP indexes.

@@

To solve the optimization problem embedded in the PPGT with JSO, a starting
projection, an index function, the number of jellyfish, and the maximum
number of iterations are provided as input. Then, the current
projection is evaluated by the index function. The projection is then
moved in a direction determined by a random factor, influenced by how
far along we are in the optimization process. Occasionally, completely
new directions may be taken like a jellyfish might with ocean currents.
A new projection is accepted if it is an improvement compared to the
current one, rejected otherwise. This process continues and iteratively
improves the projection, until the pre-specified maximum number of
trials is reached.

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of iterations

@@

This set of simulations is expanded to assess the performance of JSO
relative to the index properties, with a second data set type and
additional PP indexes. The second data set has a sine wave in two
dimensions and Gaussian noise in the remaining dimensions. Seven
additional PP indexes that should be able to extract the sine wave are
used: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `splines2d`, `skinny` and `stringy2`,
following @laa_using_2020. (The supplementary material includes
definitions of these PP indexes, and descriptions of the pipe and sine
wave data simulation.). The JSO hyper-parameters, number of jellyfish and
maximum iterations, are also included. This results in a total of
`r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine wave data. The
choice of 4-12 dimensions covers the range of easy to relatively
difficult for the optimizers to detect the structured projections.
<!-- A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability.-->
The same approach is used, that each experiment is repeated 50 times and success
rate is calculated at each level of the simulation design.

@@

Recall that smoothness and squintability are metrics used to characterize the index 
function for PP tasks and are invariant to the optimizer and its 
hyper-parameters. Thus simulations with different JSO hyper-parameters 
for the same PP task share the same smoothness value. In total, smoothness and
squintability metrics are calculated on the `r nrow(smoothness)` PP tasks in our simulation.

@@

A generalized linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting the success rate.
Predictors are smoothness, squintability, data dimension, and JSO
hyper-parameters. Due to the scale of the smoothness and squintability measure,
the rank of the two metrics, rather than the nominal values, are used in the model.

@@

```{r}
#| fig.height: 7.5
#| fig.width: 10
#| label: fig-proj
#| fig.align: "center"
#| fig.cap: "Visual comparison of JSO and CRS results, using optimal data projections obtained over 50 simulations of the pipe data. Rows correspond to data dimension. Columns correspond to quantiles of the index values, with 0 being the minimum, 100 being the maximum, and 50 the median. JSO achieves the better views of the pipe generally than CRS. As dimension increases both have more difficulty finding the pipe."

@@

The optimal projection is compared between JSO and CRS for the pipe data at
dimensions 6, 8, 10, and 12 in @fig-proj. The columns show the quantile of 
index value in the 50 optimal projection, with 0 being the minimum, 100 being 
the maximum, and 50 the median. The purpose is to summarize the views of the data resulting from
different optimization, and hence compare results between the two
optimizers. Generally, the JSO does a more consistent job of finding the
pipe structure clearly. As the data dimension increases, both optimizers
struggle and less clearly capture the circle shape.

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates sampled from 
the simulation for each case. As the number of jellyfish and maximum iteration allowed 
increase, the success rate also increases. For problems in lower dimensional (4,
6) search spaces, small parameter values (20 jellyfish and 50 iterations) 
are sufficient to achieve a high success rate. Many more
jellyfish (100 jellyfish and 100 iterations) are neede for finding the 
pipe structure in d = 10, and 12. This suggests
that even JSO will struggle with high dimensions, and require
substantial more computational time to have an acceptable success rate,
especially for difficult PP indexes based on scagnostics.

```{r}
#| label: fig-proportion
#| fig.width: 8
#| fig.height: 3.5
#| fig.align: "center"
#| fig.cap: "Summary of the relationship between JSO hyper-parameters on optimization success rate, using the holes index for pipe data of dimensions 4-12. Bootstrap samples show the variability in success rate. Success rate mostly plateaus by 50 jellyfish. The JSO has some difficulty finding the pipe when dimension is higher than 8 and maximum number of iteration has little effect."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

@@

res_quantile <- res_interval_df |> 
  unnest(df) |> 
  group_by(n_jellies, max_tries, case_id, d) |> 
  select(-I_max, -id) |> 
  distinct() |> 
  mutate(d = as.factor(d),
         n_jellies = ifelse(d == 10, n_jellies - 1, n_jellies),
         n_jellies = ifelse(d == 12, n_jellies + 1, n_jellies)) |> 
  rename(`max # of iteration` = max_tries)

res_quantile |> 
  ungroup() |> 
  ggplot(aes(x = n_jellies - 0.5, y = proportion, 
             group = d, color = d)) + 
  geom_segment(aes(xend = n_jellies + 0.5), size = 0.05) + 
  geom_segment(aes(x = n_jellies - 1, xend = n_jellies + 1), 
               data = res_quantile |> summarise(proportion = mean(proportion)), 
               size = 2) + 
  geom_line(aes(x = n_jellies), 
               data = res_quantile |> summarise(proportion = mean(proportion)), 
            alpha = 0.3) + 
  scale_color_brewer(palette = "Dark2", name = "dimension") +
  scale_x_continuous(breaks = c(20, 50, 100), labels = c("20", "50", "100")) +
  facet_wrap(vars(`max # of iteration`), 
             labeller = "label_both") + 
  theme_bw() +
  theme(panel.grid.minor = element_blank()) + 
  xlab("Number of jellyfish") + 
  ylab("Proportion of success")
```

@@

@tbl-smoothness-squintability presents values computed for a subset of
PP indexes on the sine wave data for dimensions 4-8. (A full table of
the parameters estimated from the Gaussian process (outputscale $\eta$,
lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and from the
scaled logistic function ($\theta_1$ to $\theta_4$) for calculating
smoothness and squintability for all PP indexes and data combinations is
included in the supplementary materials). The column $\nu$ is used as
the smoothness metric and the column $\varsigma$ is calculated as
equation (@eq-squintability-parametric) as the squintability metric.

@fig-idx-proj-dist shows a sample of traces from the indexes in
@tbl-smoothness-squintability for the 6D sine wave data. This traces the
PP index values computed for projections along an interpolated path
between a randomly generated noise projection and the optimal projection (sine wave). (Note
index values were self-standardized to range between 0 and 1, for
comparison.) The `splines` index is very smooth with a gradual increase
in value between the noise projection and the optimal projection. The
`TIC` index is slightly less smooth and slightly less gradual change in
value. Least desirable is the `skinny` index which is noisier and the
optimization needs to be much closer to the optimal angle to have a
higher value. This illustrates how the new measures for squintability
and smoothness work to describe the PP indexes.

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: The columns $\nu$ and $\varsigma$ show the smoothness and squintability measures for three PP indexes for the sine wave data. The splines index has the largest squint angle and is the smoothest. The skinny index has low smoothness and smallest squint angle. As dimension increases the squint angle gets smaller but smoothness doesn't change much.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 18)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = as.character(index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  filter(shape == "sine", index %in% c("TIC", "splines", "skinny")) |>
  select(shape, index, d, smooth, squint) |>
  arrange(desc(squint)) |>
  knitr::kable(
    digits = 4, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      #"$\\eta$", "$\\ell$", 
      "$\\nu$", #"$\\sigma$",
      #"$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", 
      "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rr") #align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  #kable_styling(font_size = 10)  
```

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum number of
iteration, as predictors. The fit suggests that the JSO success rate is only
affected by squintability, dimension and number of jellyfish. As expected,
the JSO success rate is higher with higher squintability and/or more
jellyfish, and lower when dimension is higher (higher-dimensional
optimization is more difficult). Interestingly, the JSO is unaffected by
the smoothness of the index function. This is consistent with the way
random search algorithms jump from value to value without taking local
regularity into account, as opposed to gradient-based optimizers for
example. Allowing for more iterations also does not affect success rate
significantly. A unit increase in squintability increases the success
rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As
dimension increases by one the success rate almost halves. Increasing
the number of jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.
@tbl-joint-output shows the fit statistics supporting that the model is
reasonably strong.

Note that the addition of interaction terms for squintability and data
dimension, and smoothness and data dimension, improves the model a small
amount, but does not change the interpretation above. This study intends 
to serve as an illustration that the proposed
measures effectively captures the key factors influencing the success
rate of PP.

```{r}
#| tbl-pos: H
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, number of jellyfish and maximum number of iterations. Interestingly, squintability and dimension strongly affect jellyfish optimization success. The number of jellies marginally affects success, but index smoothness, and increasing the number of iterations do not."
#| fig-pos: t
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate = TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of iterations")) |>
  bold_p(t = 0.1) 
```

@@

@@

Using the JSO optimizer for PPGT in the `tourr` package requires a change in user behavior. For
all the existing optimizers a single optimization path is followed. The
JSO has multiple optimization paths, and needs additional tools to
incorporate into the PPGT, as explained here.

@@

To use JSO for PPGT in the `tourr` package, specify
`search_f = search_jellyfish` in the guided tour.

The animation function `animate_*()` won't directly render the animated 
projection for JSO, as with other existing optimizers since multiple tour 
paths will need to be generated. To
visualize the path visited by each individual jellyfish, assign the
animation to an object (`res <- animate_xy(...)`). This will save the
bases visited by JSO, along with relevant metadata, as a tibble data
object [see @RJ-2021-105 for more details on the data object]. The bases
visited for individual jellyfish can then be extracted and viewed using
the `planned_tour()` function.

@@

The `ferrn` package [@RJ-2021-105] has now included functionality for computing
the smoothness and squintability metrics. Both metrics require sampling
random bases using the `sample_bases()` function, followed by the
calculation with `calc_smoothness()` or `calc_squintability()`.

@@

To sample bases for calculating smoothness, the following inputs are
required: the index function, the dataset, and the number of random
bases to sample. The output of `sample_bases()` is a data frame with a
list-column of sampled bases and index values. Parallelization
is available to speed up the index value calculation through the
`parallel = TRUE` argument.

The `calc_smoothness()` function takes the output from `sample_bases()`
and fits a Gaussian process model to the index values against the
sampled bases, as the location, to obtain the smoothness metric.
Starting parameters and additional Gaussian process arguments can be
specified and details of the fit can be accessed through the `fit_res`
attribute of the output.

@@

Bases sampling for calculating squintability includes an additional step of
interpolating between the sampled bases and the optimal basis. This step is
performed when the arguments `step_size` and `min_proj_dist` in the
`sample_bases()` function are set to non-NA numerical values. Given the
projection distance typically ranging from 0 to 2, it is recommended to
set `step_size` to 0.1 or lower, and `min_proj_dist` to be at least 0.5
to ensure a meaningful interpolation length.

The `calc_squintability()` function computes squintability using two
methods: 1) parametrically, by fitting a scaled sigmoid function through
non-linear least square (`method = "nls"`), and 2) non-parametrically,
using kernel smoothing (`method = "ks"`). A `bin_width` argument is
required to average the index values over the projection distance before
the fitting. For the parametric case, the output provides the estimated
parameters for the logistic function ($\theta_1$ to $\theta_4$)
and the calculated squintability metric as Equation
(@eq-squintability-parametric). For the non-parametric case, it shows
the maximum gradient attained (`max_d`), the corresponding projection
distance (`max_dist`), and the squintability metric as their products.

@@

This paper has presented new metrics to mathematically define desirable
features of PP indexes, squintability and smoothness, and used these to
assess the performance of the new JSO. The
metrics will be generally useful for characterizing PP indexes, and help
with developing new indexes.

@@

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optimum when more jellyfish are used and
the maximum number of iteration is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optimum. However, interestingly smoothness does not affect
the JSO performance.

Future work can focus on developing more efficient optimization algorithm for
noise indexes with low squintability metric, which resemble a needle-in-a-haystack 
problem. Additionally, if the index function is differentiable, automatic 
differentiation can be used to implement new gradient-based optimizers.
